Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-05-21 paper.Rnw
Correlations on the ADNI1 volumes figure, subregions figure

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }

%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
ADNI1:Complete 1Yr 1.5T} standardized dataset. 23 subjects were chosen from each
disease category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using a
semiautomatic tool (SNT, Medtronic Surgical Navigation Technologies, Louisville,
CO; see Supplementary Materials).

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 4: Application to the segmentation of Alzheimer's
disease patients}

@@

\paragraph{ADNI1 dataset revisited} 
All images from the {\em ADNI1:Complete 1Yr 1.5T} standardized dataset described
in Experiment 1 are used.  Clinical and demographic data are shown in table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the ADN11 Complete 1Yr 1.5T dataset are
segmented. The optimal size of template library, registration method, and label
fusion method found in Experiment 1 are used. The template library is composed
of equal numbers of images from each disease class (AD, MCI, and cognitively
normal controls).

@@

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup">>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
casted = dcast(SZ_volumes, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Manual left / right hippocampus volume " (mm^3))) + 
  ylab(expression("MAGeT left / right hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%                           RESULTS                                   %%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 4 Results: Application to the segmentation of Alzheimer's
disease patients}

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire ADNI1 Complete 1Yr 1.5T dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of
\Sexpr{length(totals.complete$RID)} images; Table 
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals}).

@@

\begin{figure}[h]
<<ADNI-scr-volumes-boxplot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
qplot(DX,value,data=melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT")),
      colour=variable,geom="boxplot") + 
      #geom_text(aes(y=value,x=DX, vjust=-1.4, size=1,
      #              label=paste("r==", signif(corr, digits=3))), 
      #      colour='black',parse=TRUE, data=snt.tests) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Total hippocampal volume (", mm^3, ")")))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-scr-volumes-plot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

snt.tests <- ddply(totals.complete, c("DX"), function(df) {
  data.frame( corr = cor(df$SNT, df$MAGeT),
              value = -Inf) 
})

ggplot(data=melted, aes(x=SNT, y=volume, colour=method)) + 
      geom_point() + 
      geom_smooth(method="lm") + 
      #geom_text(aes(y=value,x=DX, vjust=-1.4, size=1,
      #              label=paste("r==", signif(corr, digits=3))), 
      #      colour='black',parse=TRUE, data=snt.tests) +
      xlab(expression(paste("Manual total hippocampal volume (", mm^3, ")"))) + 
      ylab(expression(paste("Total hippocampal volume (", mm^3, ")")))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{fig:ADNI-scr-volumes-plot}
\end{figure}

<<ADNI-scr-volumes-table,dependson='ADNI-baseline-volumes-prep',cache=T,results='asis'>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

snt.tests <- ddply(totals.complete, c("DX"), function(df) {
  round(data.frame( MAGeT = cor(df$SNT, df$MAGeT),
              MAPER = cor(df$SNT, df$MAGeT),
              FS    = cor(df$SNT, df$FS),
              FSL   = cor(df$SNT, df$FSL)
              ), 2)
})
caption = paste(
  "Pearson correlation between manual (SNT) whole hippocampus volume", 
  "and whole hippocampus volume by method.", sep=" ")
latex(snt.tests, file="", 
      size="scriptsize",
      caption=caption, row.names="",
      title="")
@

@@

ggplot(melted, aes(x= SNT, y = diff)) + 
  geom_hline(yintercept = limits, linetype=2, color='brown') + 
 annotate("text", y = limits, x=Inf, label=round(limits,1), 
            hjust=1.1,vjust=-0.8, size=3) + 
  geom_point() + 
  xlab('Mean manual and MAGeT Brain total Hippocampus volume') + 
  ylab('Manual - MAGeT Brain volume') + 
  geom_smooth(method="lm", formula = y~x)
@
  \caption{Bland-Altman plot comparing manual (SNT) and \mb hippocampal volumes 
  in the ADNI1 1.5T 1Yr Complete dataset.}
  \label{fig:ADNI-scr-Bland-Altman}
\end{figure}

>>>>

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each
disease category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using a
semiautomatic tool (SNT, Medtronic Surgical Navigation Technologies, Louisville,
CO; see Supplementary Materials).

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "\adnidataset"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 4: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 4: Application to the segmentation of Alzheimer's
disease patients}

@@

\paragraph{ADNI1 dataset revisited} 
All images from the {\em \adnidataset} standardized dataset described
in Experiment 1 are used.  Clinical and demographic data are shown in table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the \adnidataset dataset are
segmented. The optimal size of template library, registration method, and label
fusion method found in Experiment 1 are used. The template library is composed
of equal numbers of images from each disease class (AD, MCI, and cognitively
normal controls).

@@

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup">>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
casted = dcast(SZ_volumes, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Manual unilateral whole hippocampus volume " (mm^3))) + 
  ylab(expression("MAGeT unilateral whole hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%                           RESULTS                                   %%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 4: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 4 Results: Application to the segmentation of Alzheimer's
disease patients}

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire \adnidataset dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of
\Sexpr{length(totals.complete$RID)} images; Table 
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals}).

@@

\begin{figure}[h]
<<ADNI-scr-volumes-boxplot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
qplot(DX,value,data=melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT")),
      colour=variable,geom="boxplot") + 
      #geom_text(aes(y=value,x=DX, vjust=-1.4, size=1,
      #              label=paste("r==", signif(corr, digits=3))), 
      #      colour='black',parse=TRUE, data=snt.tests) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Bilateral whole hippocampal volume (", mm^3, ")")))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-scr-volumes-plot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "Method", value.name = "volume")

df = totals.complete
levels(melted) = c("FS" = ),
                   "FSL"= paste("FSL", round(cor(df$SNT, df$FSL),2)),
                   "MAPER"= paste("MAPER", round(cor(df$SNT, df$MAPER),2)),
                   "MAGeT"= paste("MAGeT", round(cor(df$SNT, df$FS),2)))

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point() + 
    geom_smooth(method="lm") + 
    xlab(expression("Manual bilateral whole hippocampal volume " (mm^3))) + 
    ylab(expression("Automated bilateral whole hippocampal volume " (mm^3))) +
    annotate("text", y = c(3000,2500,2000,1500), x=Inf-100, size=3,
              label = c(paste("FS r=", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r=", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r=", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r=", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{fig:ADNI-scr-volumes-plot}
\end{figure}

@@

ggplot(melted, aes(x= SNT, y = diff)) + 
  geom_hline(yintercept = limits, linetype=2, color='brown') + 
  annotate("text", y = limits, x=Inf, label=round(limits,1), 
            hjust=1.1,vjust=-0.8, size=3) + 
  geom_point() + 
  xlab('Mean of manual and MAGeT-Brain hippocampal volume') + 
  ylab('Difference between Manual and MAGeT-Brain hippocampal volume') + 
  geom_smooth(method="lm", formula = y~x)
@
  \caption{Bland-Altman plot comparing manual (SNT) and \mb hippocampal volumes 
  in the \adnidataset dataset.}
  \label{fig:ADNI-scr-Bland-Altman}
\end{figure}

\begin{figure}
<<ADN1-src-subfield, cache=TRUE, dependson='ADNI-baseline-volumes-prep', fig.width=4, fig.height=4>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")

# Get subjects, and diagnoses
base = read.csv("data/ADNI_baseline_volumes/ADNI1_Screening_1.5T_11_15_2012.csv")
base = subset(base, !grepl("Scaled_2", Description)) 
base$VISCODE <- factor(base$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
subjects = yr1[,c("Image.Data.ID", "RID","VISCODE")]  # we only need these fields
dx = unique(adnimerge[c("DX.bl", "RID", "VISCODE")])
subjects = merge(subjects, dx)

# Now create one data.frame with all the yr1 volume 
# data we have from each datasource
subregions = merge(subjects, mb, by.x="Image.Data.ID", by.y="SourceImageID", 
                   all.x=TRUE)
subregions = subset(subregions, VISCODE == "bl" & RID %in% adni_69_RIDs)
attach(subregions)
subregions$X1 = mean(X1, X101)
subregions$X2 = mean(X2, X102)
subregions$X4 = mean(X4, X104)
subregions$X5 = mean(X5, X105)
subregions$X6 = mean(X6, X106)
detach(subregions)
subregions = subregions[,!names(subregions) %in% c("Left_Hippo","Right_Hippo",
                                                   "X101","X102","X103","X105",
                                                   "X106")]
subregions.melted = melt(subregions, measure.vars=c("X1","X2","X4","X5","X6"),
                         variable.name="Subregion", value.name="Volume")
levels(subregions.melted$Subregion) <-list("CA1"="X1", "CA2/CA3"="X5",
                                           "CA4/DG"="X4", 
                                           "SR/SL/SM"="X6",
                                           "Subiculum"="X2")
ggplot(subregions.melted, aes(x=Subregion, y=Volume)) +
  geom_boxplot()
@
  \caption{}
  \label{fig:ADNI-scr-subfields}
\end{figure}
\FloatBarrier
